Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2017-03-08 paper/paper-v3.tex
revision up to Analysis

%%% MF's commands
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\card}[1]{\left \lvert \, #1 \, \right\rvert}
\newcommand{\abs}[1]{\lvert #1 \rvert}
\newcommand{\States}{\ensuremath{S}\xspace}		% Set of States
\newcommand{\state}{\ensuremath{s}\xspace}		% single states
\newcommand{\mystate}[1]{\ensuremath{\state_{\text{#1}}}\xspace} %meaningful states
\newcommand{\Messgs}{\ensuremath{M}\xspace}		% Set of Messages
\newcommand{\messg}{\ensuremath{m}\xspace}		% single messages
\newcommand{\mymessg}[1]{\ensuremath{\messg_{\text{#1}}}\xspace} %meaningful messages
\newcommand{\ssome}{\mystate{\ensuremath{\exists\neg\forall}}}
\newcommand{\sall}{\mystate{\ensuremath{\forall}}}
\newcommand{\snone}{\mystate{\ensuremath{\emptyset}}}
\newcommand{\msome}{\mymessg{some}}
\newcommand{\mall}{\mymessg{all}}
\newcommand{\asome}{\myact{\ensuremath{\exists\neg\forall}}}
\newcommand{\aall}{\myact{\ensuremath{\forall}}}
\definecolor{mygray}{cmyk}{0.35,0.35,0.35,0.35}
\newcommand{\mygray}[1]{{\textcolor{mygray}{#1}}}
%%%

@@

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of a
  conventional semantic meaning of the used expression and general pragmatic reasoning applied
  to the context of utterance. This implies that models of the cultural evolution of meaning
  should likewise take into consideration that observable language use is a complex interaction
  of semantic representations and pragmatic use. To this end, we present a game theoretic model
  of the cultural evolution of language where communicative pressures work on abstract semantic
  representations and pragmatic patterns of use. Our model traces two evolutionary forces and
  their interaction: (i) fitness-based pressure towards communicative efficiency and (ii)
  systematic transmission perturbations when linguistic knowledge is transferred from one agent
  to another. The latter can arise from general cognitive or learning biases, but also from
  other sources of systematic noise, e.g., as a result of errors in perception. We illustrate
  the model based on a case study showing that cognitive biases that favor simple semantic
  representations can prevent the lexicalization of pragmatic inferences. We also show that,
  technically speaking, it is possible that environmental factors, such as perceptual errors
  during acquisition, can produce evolutionary outcomes that look as if such cognitive biases
  are present even if they are not.
\end{abstract}

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely
veiled by uttering ``I'm cold.'' The temporal succession of events can be communicated by the
order in which conjuncts appear as in ``I traveled to Paris and got married.'' An invitation
can be declined by saying ``I have to work.'' An influential explanation of the relation
between the literal meaning of expressions and what they are intended and interpreted to convey
is due to \citet{grice:1975}. It characterizes pragmatic use and interpretation as a process of
mutual reasoning about rational language use. For instance, under the assumption that the
speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
reason why the speaker will not be able to accept the invitation, going beyond its literal
meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
such as the use of ability questions for polite requests (``Could you please \dots?''), or
certain enrichments of lexical meanings such as \emph{and} to mean \emph{and then}.

A particularly productive and well studied class of pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
an utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker had them all, he
could have used the stronger word \emph{all} instead of \emph{some} in his utterance, and
thereby would have made a more informative statement. Scalar implicatures, especially the
inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
theoretically
\citep[e.g.][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many interesting details, a position endorsed by a
clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular 
process with roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved. Models of language evolution abound. There are
simulation-based models studying the evolution of language in populations of communicating
agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
work has focused on explaining basic properties such as compositionality and combinatoriality
\citep[e.g.][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
little attention has been paid to the interaction between conventional meaning and pragmatic
use. What is more, many mathematical models explain evolved meaning as a regularity in the
behavior of agents which maps objective states of the world to observable signals. There is no
room in such a purely extensional approach to address the semantics-pragmatics division
directly. Instead, we would need to look at richer representations of cognizing agents and their
communicative interaction.

To fill this gap, we spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using state-of-the-art
probabilistic cognitive models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases or systematic perceptual errors. Considering
transmission of linguistic knowledge is important because neither semantic meanings nor
pragmatic usage patterns are directly observable. Instead, language learners have to infer these
unobservables from the observable behavior in which they result. We formalize this 
process as a form of Bayesian inference. Our approach thereby contains a well-understood model
of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
with functional selection, here formalized as the most versatile dynamic from evolutionary game
theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
can lead to this outcome. Additionally, we show that a similar outcome can be reached 
without assuming any cognitive biases, but simply as an epiphenomenon of systematic disturbances from
environmental factors. This formal results highlights the frequently overlooked possibility
that channel noise in evolutionary replication can mimic effects of inductive biases.

We see the main contribution of this work as conceptual and technical, not as a definite answer
to the question why scalar implicatures emerge. The work here rather demonstrates how current
probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
combined to study the co-evolution of semantics and pragmatics side-by-side. Reversely, the
approach taken here may be seen as a first step towards giving an evolutionary rationale for
empirically successful probabilistic models of language use that embrace the majority view of
the division of labor between semantics and pragmatics. Section~\ref{sec:discussion} elaborates
on these points.

@@

The emergence and change of linguistic structure is influenced by many intertwined
factors. These range from biological and socio-ecological to cultural ones \citep{benz+etal:2005b,steels:2011,tamariz+kirby:2016}. Social and ecological pressures determine communicative needs, while
biology determines the architecture that enables and constrains the means by which they can be
fulfilled. In the following, our focus lies on the cultural aspects, wherein processes of
linguistic change are viewed as shaped by language use and its transmission, i.e., as a result
of a process of cultural evolution
\citep{Pagel2009:Human-Language-,ThompsonKirby2016:Culture-Shapes-}.

The idea that language is an adaptation to serve a communicative function has played a pivotal
role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences (e.g. in
\citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}). If processes of selection, such as conditional imitation
or reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another as natural languages need to be learnable to survive their
faithful transmission across generations. Clearly, an unlearnable code will not make it past
the one happy fellow who invented it. Importantly, even small biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process
of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. The opposition becomes particularly clear when
considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
language that associates a distinct form with all possible meanings a speaker may want to
convey is maximally expressive but challenging to acquire.

An elegant formal approach to capturing the interaction between expressivity and learnability
is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii) the
probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of an agent
with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
and its subsequent multiplication with a mutation matrix $Q$:

\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
selection in which the relative frequency of type $i$ will increase with a gradient
proportional to its average fitness in the population. This dynamic is popular and
versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
iteration of the transmission bias encoded in $Q$. In this way, the process in
(\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
iterated learning \citep{griffiths+kalish:2007}. \mf{should we include a simple example here? I 
have an example from a lecture ready at hand; it's a simple coordination game in a
one-population setting.} \tb{Yes, I think most readers would appreciate it. We can always cut it out again if the reviewers prefer more compression.}

@@

Types are what cultural evolution operates on. In standard applications of evolutionary game
theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
  one?} \tb{Yes!} For our purposes here, types are identified by their cognitive make-up. Since we are
interested in the question under which conditions processes of cultural evolution will favor
specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
consisting of a lexicon and a pragmatic strategy.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}. Lexica for this case would assign a
truth value for each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $L_4$ and the widely assumed logical
semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic
  names?} \tb{``bound'' and ``lack''? They would need to be pretty short to not be too disruptive as an inline indices.}
\begin{align*}
  L_{4} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  L_{5} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

>>>>

%%% MF's commands
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\card}[1]{\left \lvert \, #1 \, \right\rvert}
\newcommand{\abs}[1]{\lvert #1 \rvert}
\newcommand{\States}{\ensuremath{S}\xspace}		% Set of States
\newcommand{\state}{\ensuremath{s}\xspace}		% single states
\newcommand{\mystate}[1]{\ensuremath{\state_{\text{#1}}}\xspace} %meaningful states
\newcommand{\mylang}[1]{\ensuremath{L_{\text{#1}}}\xspace} %meaningful states
\newcommand{\Messgs}{\ensuremath{M}\xspace}		% Set of Messages
\newcommand{\messg}{\ensuremath{m}\xspace}		% single messages
\newcommand{\mymessg}[1]{\ensuremath{\messg_{\text{#1}}}\xspace} %meaningful messages
\newcommand{\ssome}{\mystate{\ensuremath{\exists\neg\forall}}}
\newcommand{\sall}{\mystate{\ensuremath{\forall}}}
\newcommand{\snone}{\mystate{\ensuremath{\emptyset}}}
\newcommand{\msome}{\mymessg{some}}
\newcommand{\mall}{\mymessg{all}}
\newcommand{\mnone}{\mymessg{none}}
\newcommand{\Lbound}{\mylang{bound}}
\newcommand{\Llack}{\mylang{lack}}
\newcommand{\asome}{\myact{\ensuremath{\exists\neg\forall}}}
\newcommand{\aall}{\myact{\ensuremath{\forall}}}

\definecolor{mygray}{cmyk}{0.35,0.35,0.35,0.35}
\newcommand{\mygray}[1]{{\textcolor{mygray}{#1}}}
%%%

@@

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of 
  conventional semantic meaning and general pragmatic reasoning applied
  to the context of utterance. This implies that models of cultural evolution of meaning
  should likewise take into consideration that observable language use is a complex interaction
  of semantic representations and pragmatic use. To this end, we present a game theoretic model
  of cultural language evolution where communicative pressures work on abstract semantic
  representations and pragmatic patterns of use. Our model traces two evolutionary forces and
  their interaction: (i) fitness-based pressure towards communicative efficiency and (ii)
  systematic transmission perturbations when linguistic knowledge is transferred from one agent
  to another.  We illustrate the model based on a case study showing that learning biases that favor simple semantic
  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a potential loss in expressivity from such semantics. 
\end{abstract}

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely
veiled by uttering ``I'm cold.'' The temporal succession of events can be communicated by the
order in which conjuncts appear as in ``I traveled to Paris and got married.'' An invitation
can be declined by saying ``I have to work.'' An influential explanation of the relation
between the literal meaning of expressions and what they are intended and interpreted to convey
is due to \citet{grice:1975}, characterizing pragmatic use and interpretation as a process of
mutual reasoning about rational language use. For instance, under the assumption that the
speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
reason why the speaker will not be able to accept the invitation, going beyond its literal
meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
such as the use of ability questions for polite requests (``Could you please \dots?''), or
certain enrichments of lexical meanings such as \emph{and} to mean \emph{and then}.

A particularly productive and well studied class of pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker had them all, he
could have used the stronger word \emph{all} instead of \emph{some} in his utterance, 
thereby making a more informative statement. Scalar implicatures, especially the
inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many interesting details, a position endorsed by a
clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular 
process with roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved. Models of language evolution abound. There are
simulation-based models studying the evolution of language in populations of communicating
agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much of this work has focused on explaining basic properties such as compositionality and combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
little attention has been paid to the interaction between conventional meaning and pragmatic
use. What is more, many mathematical models explain evolved meaning as a regularity in the
behavior of agents which maps objective states of the world to observable signals. There is no
room in such a purely extensional approach to address the semantics-pragmatics division
directly. Instead, we would need to look at richer representations of cognizing agents and their
communicative interaction.

To fill this gap, we spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using state-of-the-art
probabilistic cognitive models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases and general aspects of learnability. Considering
transmission of linguistic knowledge is important because neither semantic meanings nor
pragmatic usage patterns are directly observable. Instead, language learners have to infer these
unobservables from the observable behavior in which they result. We formalize this 
process as a form of Bayesian inference. Our approach thereby contains a well-understood model
of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
with functional selection, here formalized as the most versatile dynamic from evolutionary game
theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
can lead to this outcome.

We see the main contribution of this work as conceptual and technical, not as a definite answer
to the question why scalar implicatures emerge. It rather demonstrates how current
probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
combined to study the co-evolution of semantics and pragmatics side-by-side. Reversely, the
approach taken here may be seen as a first step towards giving an evolutionary rationale for
empirically successful probabilistic models of language use that embrace the majority view of
the division of labor between semantics and pragmatics. Section~\ref{sec:discussion} elaborates
on these points.

@@

The emergence and change of linguistic structure is influenced by many intertwined
factors. These range from biological and socio-ecological to cultural ones \citep{benz+etal:2005b,steels:2011,tamariz+kirby:2016}. Social and ecological pressures determine communicative needs, while
biology determines the architecture that enables and constrains the means by which they can be
fulfilled. In the following, our focus lies on cultural aspects, wherein processes of
linguistic change are viewed as shaped by language use and its transmission, i.e., as a result
of a process of cultural evolution
\citep{Pagel2009:Human-Language-,ThompsonKirby2016:Culture-Shapes-}.

The idea that language is an adaptation to serve a communicative function has played a pivotal
role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g., in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation
or reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Clearly, an unlearnable code will not make it past
the one happy fellow who invented it. Importantly, even small biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process
of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. Their opposition becomes particularly clear when
considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
language that associates a distinct form with all possible meanings a speaker may want to
convey is maximally expressive but challenging to acquire.

An elegant formal approach to capturing the interaction between expressivity and learnability
is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the step, (ii) the fitness $f_i$ of each type $i$ before the step, and (iii) the
probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of an agent
with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
and its subsequent multiplication with a mutation matrix $Q$:

\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
selection in which the relative frequency of type $i$ will increase with a gradient
proportional to its average fitness in the population. This dynamic is popular and
versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
iteration of the transmission bias encoded in $Q$. In this way, the process in
(\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
iterated learning \citep{griffiths+kalish:2007}. \mf{should we include a simple example here? I 
have an example from a lecture ready at hand; it's a simple coordination game in a
one-population setting.} \tb{Yes, I think most readers would appreciate it. We can always cut it out again if the reviewers prefer more compression.}

@@

Types are what cultural evolution operates on. In standard applications of evolutionary game
theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
  one?} \tb{Yes!} For our present purpose, types are identified by their cognitive make-up. Since we are
interested in the question under which conditions processes of cultural evolution will favor
specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
consisting of a lexicon and a pragmatic strategy.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}. \tb{I left $\snone$/$\mnone$ out because they are not relevant to the strengthening of $\msome$, but that means that it needs to be introduced a bit later. Feel free to add it if you think that's better, exposition-wise.} 
Lexica for this case would assign a
truth value for each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
semantics with only a lower-bound in $\Llack$.

\begin{align*}
  \Lbound & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  \Llack & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level-$n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?} \tb{I think that's fine. $S_n(\cdot \mid \cdot)$ is not used beyond this section and the set $S$ only by mention of its members. Otherwise we can, e.g., relabel $M$ as $F$(orms) and use $M$(eanings) instead of $S$(tates). I'm fine with either alternative.}
\begin{flalign}
&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) \propto \exp(\lambda \; L_{sm}) \label{s:level0}
\end{flalign}

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $\Lbound$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $\Llack$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; \Llack) = 0.5$. As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that higher values lead to behavior
that is increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning. Put differently, how likely the hearer is thought to infer $\state$ when sent $\messg$. The formulation
in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
necessary to guarantee a mutation matrix with only positive entries (see below). If $\lambda = 1$ a literal
speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
\sall;\Llack) \approx .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) \approx .5$ for $\Llack$.

Pragmatic behavior of level-$n+1$ is similar to its literal counterparts but uses the
interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{flalign}
